---
title: "Freezing days analysis"
author: "Ignacio Calderon, USA TODAY"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: zenburn
---

```{r setup, include=FALSE}

library(tidyverse)
library(tidycensus)
library(sf)
library(terra)
library(curl)
library(lubridate)
library(weathermetrics)
library(DT)
library(tmap)
library(svglite)
library(rmapshaper)

rm(list = ls())

knitr::opts_chunk$set(warning = F, message = F)

## check if inputs have already been downloaded

# raw climate data:
flag_dir  <- "inputs"
flag_file <- file.path(flag_dir, "input_status.txt")
noaa_exists <- file.exists(flag_file)
needs_noaa <- !noaa_exists
# processed city data:
city_data_path <- "outputs/city_data.RDS"
city_exists <- file.exists(city_data_path)
needs_city <- !city_exists
# processed county data:
county_data_path <- "outputs/county_data.RDS"
county_exists <- file.exists(county_data_path)
needs_county <- !county_exists

```

## Results

We analyzed temperature data for the past seven decades to see how winters in major cities have changed. 

We found almost all cities analyzed (255/257) are now experiencing fewer freezing days per season compared to the mid 1950s -- with 86% of cities losing at least a week of freezing days. In addition, these warming cities:

- On average are now experiencing 19 fewer freezing days per water year.
- Freezing days on average are starting 11 days later and ending 26 days earlier for a shorter cold season. 
- The longest streak of freezing days in a water year is also 12 days shorter on average.
- The coldest day of the water year has also warmed an average of 5.7 degrees Fahrenheit.

In short, the freezing season is becoming shorter and milder.

This is based on an analysis of climate data spanning 1956-2025. We extracted daily minimum temperature from the center of every city with a population over 100,000. We then ran linear regressions on those that had at least one freezing day in at least half the years.

For more details on our methods, read below.

## Methods

### Data processing

For this analysis, we used daily temperature data from [NOAA's nClimGrid-Daily](https://www.ncei.noaa.gov/products/land-based-station/nclimgrid-daily). This is gridded product has a resolution is about 5-km and goes back to 1951. 

The data can be accessed in many formats but we compiled netCDF monthly files from 1956-2025 for a complete 70 years. These 800+ files have a total size of ~50 GB. We started by automating the extraction like this:

```{r c1, eval = needs_noaa}
# first let's make the URLs to extract the raw data for 1955:2025
# adding a year before to complete winter season for first year

years <- 1955:2025
months <- sprintf("%02d", 1:12)
base <- "https://www.ncei.noaa.gov/data/nclimgrid-daily/access/grids/"

# create url combinations
links <- expand.grid(year = years, month = months) %>%
  mutate(url = paste0(base, year, "/ncdd-", year, month, "-grd-scaled.nc")) %>%
  arrange(year)

# now create a function that will download the files and store them in lists
download_fc <- function(var){
  
  # define the target directory and create it if it doesn't exist
  dir.create("inputs/noaa", showWarnings = F)
  
  # define the file path for the downloaded file
  file_path <- file.path("inputs/noaa", basename(var))
  
  # download the file
  curl_download(var, file_path)
  
  # read the netCDF file
  data <- rast(paste0("inputs/noaa/", basename(var)))
  
  return(data)
  
}

# iterate this for both variables
data_list <- lapply(links$url, download_fc)

rm(links, base, months, years)

# ---- status write
# create a txt file with metadata on the data download
# that way next time we run the script and input data is present, we can skip this chunk

# define where the data and flag live
data_root <- "inputs/noaa"

# get a quick summary
all_files   <- list.files(data_root, recursive = TRUE, full.names = TRUE)
total_bytes <- sum(file.info(all_files)$size, na.rm = TRUE)

# write to a temp file and then atomically rename to avoid partial/corrupt flags
tmp_flag <- tempfile(pattern = "input_status_", tmpdir = flag_dir)

status_lines <- c(
  sprintf("downloaded = TRUE"),
  sprintf("timestamp = %s", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
  sprintf("data_root = %s", data_root),
  sprintf("file_count = %d", length(all_files)),
  sprintf("total_bytes = %s", format(total_bytes, big.mark = ",")))

writeLines(status_lines, con = tmp_flag)
file.rename(tmp_flag, flag_file)

rm(all_files, total_bytes, tmp_flag, status_lines, data_root, flag_dir, flag_file)

```

```{r c1B, eval = noaa_exists}
## if the input files are already downloaded, import them here instead:

# list the files
data_files <- list.files("inputs/noaa", full.names = TRUE)

# load the files as SpatRaster objects
data_list <- lapply(data_files, rast)

rm(data_files)

# check units, projections, number of layers for a sample monthly file
# data_list[[1]] # Celsius
# st_crs(data_list[[1]]) # WGS 84
# nlyr(data_list[[1]]) # 124 layers. 1/variable/day (tmin, tmix, prcp, tavg)
```

This data has daily min, max and average temperature in degrees Celsius. Before we can use it, we need to chose the areas of interest. Since the data comes in a grid, the analysis can be done at different geographic levels. We will start with cities.

We used the `tidycensus` package to query all cities and their corresponding boundaries. We then filtered for major cities. We defined this as population greater than or equal to 100,000 and ended up with 310 cities.

```{r c2, eval = needs_city}
# first let's get a list of census "places"
# this includes cities, towns, villages
options(tigris_use_cache = TRUE)
us_places <- get_acs(geography = "place",
                     variables = "B01003_001", # total population
                     year = 2022,
                     survey = "acs5",
                     output = "wide",
                     geometry = TRUE) %>%
                     select(NAME, GEOID, pop = B01003_001E, geometry)

# can filter for cities with pop > 100k
us_cities_big <- us_places %>%
  filter(str_detect(NAME, "city")) %>%
  filter(pop >= 100000)

rm(us_places)

# clean up some names
us_cities_big <- us_cities_big %>%
  mutate(NAME = str_replace(NAME, "San Buenaventura \\(Ventura\\)", "Ventura"),
         NAME = str_replace(NAME, " \\(.*\\)", ""),
         NAME = str_replace(NAME, " city,", ","))
```

With a list of cities, we also calculated the centroids for each. This lets us convert the gridded climate data into point data. To do this, we made a unique key from the city coordinates. With that key, we can tell R where to grab a city's temperature from each day layer in the map.

The result is a database with a daily timeseries for each city. The final step before we finish prepping the raw data is converting the temperature units to Fahrenheit.

```{r c3, eval = needs_city}
# check projection
# st_crs(us_cities_big) # NAD83

# get the centroids 
us_cities_big <- us_cities_big %>%
  mutate(center = st_centroid(geometry)) %>%
  st_drop_geometry() %>% # removes city outline but keeps centroids
  st_as_sf() %>%
  st_transform(crs = st_crs(data_list[[1]])) # reproject to match our climate data

# we'll do this by getting a sample layer from temp data
# and making a key from the x and y coords
sample_layer <- data_list[[1]][[1]] %>% # first monthly file, first day
  as.data.frame(xy = TRUE) %>%
  mutate(key = paste0(round(x, 5), "_", round(y, 5))) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(data_list[[1]])) %>%
  select(-contains("tmax_"))

# if we join this to our city list,
# we'll have a key that connects each city to its nearest pixel in the climate data
key_df <- us_cities_big %>%
  st_join(sample_layer, join = st_nearest_feature) %>%
  select(NAME, GEOID, key) %>%
  st_drop_geometry()

# check how many keys were created
# length(unique(key_df$key)) # 310 keys for 310 cities

rm(sample_layer)

## now we can extract the data from the gridded file

# convert city centers to a terra SpatVector object
city_points <- vect(us_cities_big)

# let's make a function to extract temperature values
# for each city center over all time layers
extract_fc <- function(gridded_data, place_points, key_df){
  
  # extract variables from grid
  data <- terra::extract(gridded_data, place_points, xy = TRUE, ID = FALSE)
  
  # the extraction leaves one column for each day/variable plus columns for x and y coords
  # we can add city names/GEOID back using our crosswalk
  data <- data %>%
    # make key from x and y the same way we did before
    mutate(key = paste0(round(x, 5), "_", round(y, 5))) %>%
    # now join
    left_join(key_df, by = "key")
  
  # here, we can remove precipitation data since we don't need it
  # also drop tmax_ and tavg_ for now since processing can get heavy
  data <- data %>%
    select(-contains("prcp_"), -starts_with("tmax_"), -starts_with("tavg_"))
  
  # pivot longer
  data_long <- data %>%
    pivot_longer(cols = c(starts_with("tmin_")), # can add tmax_ and tavg_ here
                 names_to = c("measure", "day"),
                 names_pattern = "(tmin)_(\\d+)", # (tmax|tmin|tavg)
                 values_to = "celsius") %>%
    mutate(day = as.integer(day))
  
  # now we need to add data
  # create a lookup table for day -> date
  dates <- terra::time(gridded_data)
  
  date_lookup <- tibble(date = dates) %>%
    mutate(day = as.integer(sub(".*-", "", date))) %>%
    distinct()
  
  # join that
  data_long <- left_join(data_long, date_lookup, by = "day") %>% 
    select(place = NAME, key, GEOID, measure, celsius, date)
  
  return(data_long)
  
}

# iterate
city_data <- map_dfr(.x = data_list,
                     .f = extract_fc,
                     place_points = city_points,
                     key_df = key_df)

rm(city_points, us_cities_big)

# save this as an R object to save processing time
saveRDS(city_data, city_data_path)
```

```{r c4}

# if this is a quick run, read the input file
# if it's a full run, the data was created in the last chunk, so carry on
if (needs_city == F) {
  city_data <- readRDS(city_data_path)
} else {
  message("Using input files from previous computation")
}

# convert celsius to F
# then let's pivot wide the variables while keeping one row per day
city_data2 <- city_data %>%
  mutate(fahrenheit = celsius.to.fahrenheit(celsius),
         measure = paste0(measure, "_F")) %>%
  select(-celsius, -key) %>%
  pivot_wider(names_from = 'measure', values_from = 'fahrenheit')

rm(city_data)
```

### Analysis

One way to check for long-term changes in the colder season is by looking at the frequency and intensity of freezing days (t-minimum <= 32F). Since winters in the north hemisphere are split by our arbitrary calendar year, we can aggregate this count by water year instead. Think of it as the fiscal year used in climate and water research.

The analysis would end with water year 2025, which started in Oct. 1, 2024 and ran through Sept. 31, 2025.

These are the specific metrics we quantified for each water year:

- Number of freezing days
- Date of the earliest and latest freezing day (to calculate the freezing season length)
- Duration of longest streak of freezing days
- Lowest daily minimum for the whole water year

```{r c5}
# calculate year/water year for aggregations
# computations below get heavy so we can split by chunks
city_data2_list <- city_data2 %>%
  group_split(GEOID, .keep = TRUE)

rm(city_data2)

process_data2 <- function(df) {
  df %>%
    mutate(year = year(date),
         month = month(date),
         water_year = if_else(month >= 10, year + 1, year)) %>%
    select(-month) %>%
    mutate(water_year_start = as.Date(paste0(water_year - 1, "-10-01")),
           doy_water = as.numeric(difftime(date, water_year_start, units = "days")) + 1) %>%
    select(-water_year_start) %>%
    # remove first water year (incomplete, we got that calendar year to fill the next)
    filter(water_year >= 1956)
}

city_data3 <- map_dfr(city_data2_list, process_data2)

rm(city_data2_list)

# aggregate data by WATER year and calculate from MIN temp:
# - number of freezing days (<= 32F),
# - date for the first and last day each city reached freezing temp
# - lowest daily minimum

# split by GEOID again to process in batches
city_data3_list <- city_data3 %>%
  group_split(GEOID, .keep = TRUE)

rm(city_data3)

process_yearly_stats <- function(df) {
  df %>%
    mutate(freezing = ifelse(tmin_F <= 32, 1, 0)) %>%
    group_by(place, GEOID, water_year) %>%
    summarise(freezing_days = sum(freezing, na.rm = T),
              freezing_first = min(doy_water[freezing == 1], na.rm = TRUE),
              freezing_last = max(doy_water[freezing == 1], na.rm = TRUE),
              coldest_min = min(tmin_F, na.rm = T)) %>%
    # clean up these variables, replacing inf values for NA
    mutate(freezing_first = replace(freezing_first, is.infinite(freezing_first), NA),
           freezing_last = replace(freezing_last, is.infinite(freezing_last), NA),
           # calculate season "length" (start to end of freezing days) if both dates exist
           season_length = freezing_last - freezing_first)
}

yearly_stats <- map_dfr(city_data3_list, process_yearly_stats)

## the aggregation above will produce many warnings for places that
## don't get to below 32F but those are handled

# separately, calculate the longest freezing days streak
# by MIN temp per WATER year

process_longest <- function(df) {
  df %>%
    arrange(place, year, date) %>% # ensure proper ordering
    group_by(place, GEOID, water_year) %>%
    mutate(freezing = ifelse(tmin_F <= 32, 1, 0),  # binary column
           streak_id = cumsum(!freezing)) %>%  # assigns unique IDs to streaks
    filter(freezing == 1) %>%  # keep only rows where avg temp <= 0
    # record the longest for each city/water year
    summarise(freezing_longest_streak = max(rle(streak_id)$lengths, na.rm = TRUE), .groups = "drop") 
}

longest <- map_dfr(city_data3_list, process_longest)

# now join this back to yearly stats
yearly_stats <- left_join(yearly_stats, longest)
rm(longest, city_data3_list)

```

With a water-year aggregation by city, we can run individual metrics through linear regressions. While there are multiple stat tools to check the trend, this gives us a simple number, the slope, that describes: if the annual number of freezing days in a given place is increasing or decreasing (positive or negative value) and how quickly that change is happening.

Multiplying the slope by the number of years (70) also gives us the total change over time. This is a better way to describe the change than comparing the first and last years as those can be biased with shorter-term variations of an unusually cold/hot year.

In the last section we'll use a second stat tool to check if the trend is significant. For now, here's how we iterated the regressions. Note, to make this more robust, we only included cities that had at least one freezing day in at least half the years (n = 257).

```{r c6}
# we will only keep cities that had at least 1 freezing day 
# in at least half the years analyzed
complete <- yearly_stats %>%
  mutate(freezing_water_year = ifelse(freezing_days >= 1, 1, 0)) %>%
  group_by(place, GEOID) %>%
  summarise(water_years_w_freezing = sum(freezing_water_year),
            pct_water_years_freezing = water_years_w_freezing/n()) %>%
  filter(pct_water_years_freezing >= 0.5)

# filter for these
yearly_stats_complete <- yearly_stats %>%
  filter(GEOID %in% complete$GEOID)

# now we'll split them in a list to apply some custom functions
yearly_list <- split(yearly_stats_complete, yearly_stats_complete$GEOID)

# to test change over time,
# we will run linear regressions
total_change_fc <- function(df) {
  
  start_year <- min(df$water_year)
  end_year <- max(df$water_year)
  
  # set all the variables we'll test
  vars <- c("freezing_days", "freezing_first", "freezing_last", 
            "coldest_min", "season_length", "freezing_longest_streak")
  
  # initialize an empty list to store results
  results <- list()
  
  # loop over variables and fit the linear model
  for (var in vars) {
    
    # if a city has not experienced any freezing days then 
    # freezing_days will be zero across all rows, and all the other vars will be NA
    # when this happens, don't run any regressions and return NA results instead
    if (sum(!is.na(df[[var]])) == 0 | sum(df$freezing_days == 0) == nrow(df)) {
      
      warning(paste("All values are NA or zero for", var, "in city:", unique(df$city)))
      results[[var]] <- NA
      
    } else {
      
      ## if there's enough valid data, run this:
      
      # regression
      model <- lm(as.formula(paste(var, "~ water_year")), data = df)
      
      # extract slope (change per water_year) and intercept
      slope <- coef(model)["water_year"]
      intercept <- coef(model)["(Intercept)"]
    
      # fitted/modeled values at the endpoints 
      # f(x) = mx + b
      start_fit <- (slope * start_year) + intercept
      end_fit   <- (slope * end_year) + intercept
      
      # calculate total change over the period
      total_change <- slope * (end_year - start_year)
      
      # store results
      results[[paste0(var, "_change")]] <- total_change
      results[[paste0(var, "_", start_year)]] <- start_fit
      results[[paste0(var, "_", end_year)]] <- end_fit

    }
  }
  
  # convert the list into a dataframe and add GEOID/place name
  results_df <- as.data.frame(results) %>%
    mutate(GEOID = unique(df$GEOID),
           place = unique(df$place)) %>%
    select(GEOID, place, everything()) %>%
    # round all numeric results to whole numbers (days),
    # except coldest_min columns which keeps one decimal
    mutate(across(where(is.numeric) & !contains("coldest_min"), round),
           across(contains("coldest_min") & where(is.numeric), ~ round(.x, 1)))

  return(results_df)
  
}

# iterate through data
# this represents the total change across the study period (1956-2025)
# also included are the modeled values for the start and end years for each metric
total_change <- lapply(yearly_list, total_change_fc) %>% bind_rows()
rownames(total_change) <- NULL

# export results both by year and aggregated
write_csv(yearly_stats, "outputs/yearly_stats_city.csv")
write_csv(total_change, "outputs/total_change_city.csv")
```

### Results

This analysis shows 99% of major cities included (255/257) experienced an decrease in the number of freezing days since 1956. From those:

- On average are now experiencing 19 fewer freezing days per water year.
- Freezing days on average are starting 11 days later and ending 26 days earlier for a shorter cold season. 
- The longest streak of freezing days in a water year is also 12 days shorter on average.
- The coldest day of the water year has also warmed an average of 5.7 degrees Fahrenheit.

When we ran the analysis to 1975-2024 the results aligned with earlier reports from [Climate Central](https://www.climatecentral.org/climate-matters/fewer-freezing-nights-2025) looking at a similar time period.

A city-level summary table of the total change over time can be found below.

Here's an example interpretation of the variables: if Fort Collins, Colorado says -10 for **freezing_days_change**, that means it now experiences 10 fewer freezing days than in the mid 1950s. A value of -67 for **season_length_change** means period of time between the first and last freezing day of the water year has gotten 67 days shorter (a combination of freezing starting later and ending sooner). The value under **cold_min_change** refers to how many degrees F is the coldest day in a water year warmed or cooled compared to the mid 1950s. Lastly, a value of -17 for **freezing_longest_streak_change** means the longest consecutive run of freezing days in a water year is now 17 days shorter than in the mid 1950s.

```{r c7}
# at city level: 99% (255 out of 257 major cities)
nrow(filter(total_change, freezing_days_change < 0))/nrow(total_change)

# from those:
# - avg decrease was 19 days/water year
# - a quarter had a decrease of 26/year
# - freezing days are starting an average of 11 days later
# and ending 26 days earlier for a season 37 days shorter
# - the coldest min warmed by an average of 5.7F 
# - even the longest streak in a season has decreased by an average of 12 days
summary(filter(total_change, freezing_days_change < 0))

# at city level: 86% lost at least 7 freezing days
# (220 out of 257 major cities)
nrow(filter(total_change, freezing_days_change <= -7))/nrow(total_change)

# at city level: 67% lost at least 14 freezing days
# (173 out of 257 major cities)
nrow(filter(total_change, freezing_days_change <= -14))/nrow(total_change)

# we can also check how these rankings by climate regions
states_info <- fips_codes %>%
  select(state_fip = state_code, state_abb = state, state_name) %>%
  distinct() %>%
  # data is only for CONUS, remove Puerto Rico and territories
  filter(!state_abb %in% c("AK", "HI", "AS", "GU", "MP", "PR", "UM", "VI")) %>%
  # from noaa: https://www.ncei.noaa.gov/access/monitoring/reference-maps/us-climate-regions
  mutate(climate_division = case_when(
    state_name %in% c("Connecticut","Delaware","Maine","Maryland",
                      "Massachusetts","New Hampshire","New Jersey",
                      "New York","Pennsylvania","Rhode Island","Vermont") ~ "Northeast",
    state_name %in% c("Iowa","Michigan","Minnesota","Wisconsin") ~ "Upper Midwest",
    state_name %in% c("Illinois","Indiana","Kentucky","Missouri",
                      "Ohio","Tennessee","West Virginia") ~ "Ohio Valley",
    state_name %in% c("Alabama","Florida","Georgia","North Carolina",
                      "South Carolina","Virginia") ~ "Southeast",
    state_name %in% c("Montana","Nebraska","North Dakota","South Dakota",
                      "Wyoming") ~ "Northern Rockies and Plains",
    state_name %in% c("Arkansas","Kansas","Louisiana","Mississippi",
                      "Oklahoma","Texas") ~ "South",
    state_name %in% c("Arizona","Colorado","New Mexico","Utah") ~ "Southwest",
    state_name %in% c("Idaho","Oregon","Washington") ~ "Northwest",
    state_name %in% c("California","Nevada") ~ "West"))

# join this to our findings
total_change <- total_change %>%
  mutate(state_fip = substr(GEOID, 1, 2)) %>%
  left_join(select(states_info, state_fip, climate_division), by = "state_fip") %>%
  select(-state_fip)

# let's look further into the which places increased
total_change_decrease <- total_change %>%
  filter(freezing_days_change < 0) %>%
  arrange(freezing_days_change)

# 9 of the top 10 decreases are in the upper midwest, ohio valley and the northeast
# head(total_change_decrease, 10)
```

```{r e1, echo = FALSE}
# display this in a table
total_change %>%
  select(city = place, freezing_days_change, coldest_min_change, season_length_change, 
         freezing_longest_streak_change, climate_division) %>%
  arrange(freezing_days_change) %>%
  datatable(extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf')))
```

While the summary stats above describe the overall change through the 70 years, we can also calculate the average number of freezing days (and the average for all the other metrics) by water year across the cities that saw a decrease in freezing days for more visualizations.

Note, not all cities had freezing days every year. If a city didn't have freezing days, a value of 0 is used for the water year average calculation, but because freezing_first and freezing_last are NA, those are removed. Because our analysis already filtered to use those that had at least 1 freezing day in at least half the years, the removals shouldn't have a big impact. A column for how many had freezing days per year is included (mean is 244 cities and median 248, so close to the full set of 255).

```{r city_average}
# open them here if not loaded
# yearly_stats <- read_csv("outputs/yearly_stats_city.csv")
# total_change <- read_csv("outputs/total_change_city.csv")

# filter cities that saw a decrease
total_change_decrease <- total_change %>%
  filter(freezing_days_change < 0)

# now average these by year
yearly_stats_avg <- yearly_stats %>%
  filter(GEOID %in% total_change_decrease$GEOID) %>%
  group_by(water_year) %>%
  summarise(n_cities_freeze = sum(replace_na(freezing_days, 0) > 0),
            mean_freezing_days = round(mean(freezing_days, na.rm = T)),
            mean_freezing_first = round(mean(freezing_first, na.rm = T)),
            mean_freezing_last = round(mean(freezing_last, na.rm = T)),
            mean_coldest_min_F = round(mean(coldest_min, na.rm = T)))

# check coverage by year, out of 255
# summary(yearly_stats_avg$n_cities_freeze)
```

```{r fever, echo = FALSE, fig.height=4, fig.width=6, fig.align = 'center'}
# line chart
breaks_decade <- c(seq(1956, 2016, by = 10), 2025)

yearly_stats_avg %>%
  ggplot(aes(x = water_year, y = mean_freezing_days)) +
  geom_line() +
  ylim(40, 90) +
  scale_x_continuous(breaks = breaks_decade, labels = breaks_decade) +
  labs(title = "Average number of freezing days by water year",
       subtitle = "Based on observed values from the 255 cities that saw an overall decrease",
       x = NULL, y = NULL) +
  theme_linedraw() +
  theme(axis.title = element_blank())
```

For another visualization that compares the start and end of our study period, we also used the average across the cities that saw a decrease in freezing days but this time using values from the regression lines to account for the shorter term variability mentioned earlier since we are only displaying two years.

```{r lollipop, echo = FALSE, fig.height=1.5, fig.width=6, fig.align = 'center'}
# for another version of the viz, let's get the average 
# start/end freezing dates from the regression values
lollipop_df <- data.frame(water_year = c(1956, 2025),
                          mean_freezing_first = c(mean(total_change_decrease$freezing_first_1956), 
                                                  mean(total_change_decrease$freezing_first_2025)),
                          mean_freezing_last = c(mean(total_change_decrease$freezing_last_1956), 
                                                 mean(total_change_decrease$freezing_last_2025))) %>%
  mutate(# reverse so oldest (1956) is on top
        water_year_f = factor(water_year, levels = rev(sort(unique(water_year)))),
        # water-year start date (Oct 1 of prior calendar year)
        wy_start = as.Date(paste0(water_year - 1, "-10-01")),
        # convert day-of-water-year to real dates
        start_date = wy_start + (mean_freezing_first - 1),
        end_date = wy_start + (mean_freezing_last - 1),
        # format to M/D
        start_lab = format(start_date, "%m/%d"),
        start_lab = gsub("^0", "", start_lab),  # drop leading 0 in month
        start_lab = gsub("/0", "/", start_lab), # drop leading 0 in day
        end_lab = format(end_date, "%m/%d"),
        end_lab = gsub("^0", "", end_lab),
        end_lab = gsub("/0", "/", end_lab))

# water year month ticks: Oct 1 = 1, Nov 1 = 32, ...
month_breaks <- c(1, 32, 62, 93, 124, 152, 183, 213)
month_labels <- c("Oct","Nov","Dec","Jan","Feb","Mar","Apr","May")

# offset (in water-year days) to place labels slightly away from the points
label_pad <- 3

ggplot(lollipop_df, aes(y = water_year_f)) +
  # segment (the "lollipop stick")
  geom_segment(aes(x = mean_freezing_first,
                   xend = mean_freezing_last,
                   yend = water_year_f),
               linewidth = 1, color = "grey55") +
  # endpoints
  geom_point(aes(x = mean_freezing_first, color = "Start"), size = 2) +
  geom_point(aes(x = mean_freezing_last,  color = "End"),   size = 2) +
  # left label for start (to the left of the start tip)
  geom_text(aes(x = mean_freezing_first - label_pad, label = start_lab),
            hjust = 1, vjust = 0.5, size = 3, color = "grey20") +
  # right label for end (to the right of the end tip)
  geom_text(aes(x = mean_freezing_last + label_pad, label = end_lab),
            hjust = 0, vjust = 0.5, size = 3, color = "grey20") +
  scale_x_continuous(limits = c(1, 213),
                     breaks = month_breaks,
                     labels = month_labels,
                     expand = c(0.01, 0.01)) +
  labs(title = "Average start and end of freezing days by water year",
       subtitle = "Using linear regression values",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title = element_blank())

```

### Validation

To check our results in a different way, we can use a paired t-test. If we calculate the average annual freezing days over the two most recent decades (2006-2025) at a given location, we can pair this data point with the average for the first two decades (1956-1975) in the same place. Then, we can run this test across all cities with complete data, including those that are getting colder. We repeated this for the winter metrics we are tracking: number of freezing days, season length, longest freezing day streak, coldest temp of the year.

Since the p-values for all tests are well below 0.01, it means we can reject the null hypothesis (no change over time), to say winters in these cities are generally getting shorter and milder.

```{r c8}
# split annual data into the first 2 decades vs the past 2 decades
# aggregate our stats by the city/period to get "paired" numbers
period <- yearly_stats_complete %>%
  mutate(period = case_when(
    water_year %in% c(1956:1975) ~ "first_half",
    water_year %in% c(2006:2025) ~ "second_half")) %>%
  group_by(place, period) %>%
  summarise(avg_freezing_days = mean(freezing_days, na.rm = TRUE),
            avg_season_length = mean(season_length, na.rm = TRUE),
            avg_coldest_min = mean(coldest_min, na.rm = T),
            avg_freezing_longest_streak = mean(freezing_longest_streak, na.rm = TRUE)) %>%
  drop_na(period)
  
period_wide <- period %>%
  pivot_wider(names_from = period,
              values_from = c(avg_freezing_days, avg_season_length, avg_coldest_min, avg_freezing_longest_streak)) %>%
  arrange(place)  # explicit, so x and y match

## we can test for statistically significant changes using a t-test

# since p < 0.01, this result is statistically significant at the 1% significance level 
# we can reject the null hypothesis and conclude that there's been an decrease in
# freezing days across major US cities
t.test(period_wide$avg_freezing_days_second_half, 
       period_wide$avg_freezing_days_first_half, 
       paired = TRUE,
       alternative = "less") # less days now

# same for avg_season_length
t.test(period_wide$avg_season_length_second_half, 
       period_wide$avg_season_length_first_half, 
       paired = TRUE, 
       alternative = "less") # shorter season now

# and for avg_coldest_min
t.test(period_wide$avg_coldest_min_second_half, 
       period_wide$avg_coldest_min_first_half, 
       paired = TRUE, 
       alternative = "greater") # higher temp now

# and for avg_freezing_longest_streak
t.test(period_wide$avg_freezing_longest_streak_second_half, 
       period_wide$avg_freezing_longest_streak_first_half, 
       paired = TRUE, 
       alternative = "less") # shorter freezing streak now
```

## County-level data

For further exploration of the data (like mapping), we can also use the gridded files to extract data at a county level using the same approach as before. To spare you from the repeat code, here are the two differences:

- Use county geometry centroids (filtered by lower 48 states). There were only 13 counties where this approached failed to get data. These are places where the centroid fell on a lake or river. For these, we used the centroid of the biggest city/town within that county to ensure the centroids fell on land and data could be extracted.
- Rather than only keeping cities with at least 1 freezing day in at least half the years, we ran it to all counties to keep the data spatially "complete."

```{r c9, echo = FALSE}
# get a list of all counties
options(tigris_use_cache = TRUE)
us_counties <- get_acs(geography = "county",
                    variables = "B01003_001", # total population
                    year = 2022,
                    survey = "acs5",
                    output = "wide",
                    geometry = TRUE) %>%
  select(NAME, GEOID, geometry)

# get state FIPS to remove Hawaii, Alaska, Puerto Rico
# we will also remove some independent cities in Virginia that show up with counties
states_info <- fips_codes %>%
  select(state_fip = state_code, state_abb = state, state_name) %>%
  distinct()

us_counties <- us_counties %>%
  mutate(state_fip = substr(GEOID, 1, 2)) %>%
  left_join(states_info, by = "state_fip") %>%
  # remove AK and HI
  filter(!state_abb %in% c("AK", "HI")) %>%
  # and remove PR plus any territories
  filter(!state_abb %in% c("AS", "GU", "MP", "PR", "UM", "VI")) %>%
  # finally remove some cities in VA
  filter(!str_detect(NAME, "\\bcity\\b")) %>% # keeps places like "James City County" but not "Harrisonburg city"
  select(-state_fip, -state_abb, -state_name)

# get the centroids 
us_counties_cent <- us_counties %>%
  mutate(center = st_centroid(geometry)) %>%
  st_drop_geometry() %>% # removes city outline but keeps centroids
  st_as_sf() %>%
  st_transform(crs = st_crs(data_list[[1]])) # reproject to match our climate data

# because there are 13 counties that have centroids that fall
# on or near a body of water, those fail to extract climate data
# we will overwrite their centroids to the center of their
# most populated city/town
missing_GEOID <- c("25001", "51131", "26083", "53055", "49011", "44005", 
             "53029", "38055", "12087", "21143", "27007", "37143", "50013")

missing_states <- c("FL", "KY", "MA", "MI", "MN", "NC", "ND", "RI", "UT", "VT",
                    "VA", "WA")

us_counties_miss <- us_counties %>% filter(GEOID %in% missing_GEOID)

missing_places <- get_acs(
  geography = "place",
  variables = "B01003_001", # total population
  year = 2022,
  survey = "acs5",
  output = "wide",
  geometry = TRUE) %>%
  select(NAME_place = NAME, GEOID_place = GEOID, pop = B01003_001E, 
         geometry_place = geometry) %>%
  mutate(state_fip = substr(GEOID_place, 1, 2)) %>%
  left_join(states_info, by = "state_fip") %>%
  # only keep from states that have missing counties
  filter(state_abb %in% missing_states) %>%
  select(-state_fip, -state_abb, -state_name)

# spatial join: assign each place to a county
# now we can narrow to the county as well
places_within_county <- st_join(missing_places, us_counties_miss, 
                                join = st_within) %>% drop_na(GEOID)

# get the centroid for the largest place per county
us_counties_miss_cent <- places_within_county %>%
  group_by(NAME, GEOID) %>%
  slice_max(pop, n = 1) %>%
  ungroup() %>%
  select(NAME, GEOID, geometry = geometry_place) %>%
  mutate(center = st_centroid(geometry)) %>%
  st_drop_geometry() %>% # removes county outline but keeps centroids
  st_as_sf() %>%
  st_transform(crs = st_crs(data_list[[1]])) # reproject to match our climate data

# now replace those from the largest dataset
us_counties_cent <- us_counties_cent %>%
  filter(!GEOID %in% missing_GEOID) %>%
  rbind(us_counties_miss_cent)

rm(missing_states, missing_places, places_within_county, us_counties_miss, us_counties_miss_cent)
  
# check projection
# st_crs(us_counties) # NAD83
```

```{r c10, eval = needs_county, echo = FALSE}
# now we can get back to making a sample layer from temp data
# and making a key from the x and y coords
sample_layer_county <- data_list[[1]][[1]] %>% # first monthly file, first day
  as.data.frame(xy = TRUE) %>%
  mutate(key = paste0(round(x, 5), "_", round(y, 5))) %>%
  st_as_sf(coords = c("x", "y"), crs = st_crs(data_list[[1]])) %>%
  select(-contains("tmax_"))

# if we join this to our city list,
# we'll have a key that connects each city to its nearest pixel in the climate data
key_df_county <- us_counties_cent %>%
  st_join(sample_layer_county, join = st_nearest_feature) %>%
  select(NAME, GEOID, key) %>%
  st_drop_geometry()

# check how many keys were created
# length(unique(key_df_county$key)) # 3069 keys for 3069 counties

rm(sample_layer_county)

## now we can extract the data from the gridded file

# convert city centers to a terra SpatVector object
county_points <- vect(us_counties_cent)

# reuse the extract function
county_data <- map_dfr(.x = data_list,
                       .f = extract_fc,
                       place_points = county_points,
                       key_df = key_df_county)

rm(county_points, data_list, us_counties_cent)

# save this as an R object to save processing time
saveRDS(county_data, county_data_path)
```

```{r c11, echo = FALSE}

# if this is a quick run, read the input file
# if it's a full run, the will was created in the last chunk, so carry on
if (needs_county == F) {
  county_data <- readRDS(county_data_path)
} else {
  message("Using input files from previous computation")
}

# convert celsius to F
# then let's pivot wide the variables while keeping one row per day
county_data2 <- county_data %>%
  mutate(fahrenheit = celsius.to.fahrenheit(celsius),
         measure = paste0(measure, "_F")) %>%
  select(-celsius, -key) %>%
  pivot_wider(names_from = 'measure', values_from = 'fahrenheit')

rm(county_data)

# calculate year/water year for aggregations
# computations below get heavy so we can split by chunks
county_data2_list <- county_data2 %>%
  group_split(GEOID, .keep = TRUE)

rm(county_data2)

county_data3 <- map_dfr(county_data2_list, process_data2)

rm(county_data2_list)

# aggregate data by WATER year and calculate from MIN temp:
# - number of freezing days (<= 32F),
# - date for the first and last day each city reached freezing temp
# - lowest daily minimum

# split by GEOID again to process in batches
county_data3_list <- county_data3 %>%
  group_split(GEOID, .keep = TRUE)

rm(county_data3)

yearly_stats_county <- map_dfr(county_data3_list, process_yearly_stats)

## the aggregation above will produce many warnings for places that
## don't get to below 32F but those are handled

# separately, calculate the longest freezing days streak
# by MIN temp per WATER year
longest_county <- map_dfr(county_data3_list, process_longest)

# now join this back to yearly stats
yearly_stats_county <- left_join(yearly_stats_county, longest_county)
rm(longest_county, county_data3_list)

# now we'll split them in a list to apply some custom functions
yearly_list_county <- split(yearly_stats_county, yearly_stats_county$GEOID)

# reuse total_change_fc to iterate on county data
# this represents the total change across the study period (1956-2025)
total_change_county <- lapply(yearly_list_county, total_change_fc) %>% bind_rows()
rownames(total_change_county) <- NULL

rm(yearly_list_county)

# export results both by year and aggregated
write_csv(yearly_stats_county, "outputs/yearly_stats_county.csv")
write_csv(total_change_county, "outputs/total_change_county.csv")
```

After processing the data, we can join the results back to the geometries to map it. Here is the total change in freezing days between 1956-2025:

```{r c12, echo = FALSE, fig.align = 'center'}
# join geometries back
total_change_county_geom <- total_change_county %>%
  left_join(us_counties, by = "GEOID") %>%
  rename(county_name = NAME)

# set map to interactive, for static do: tmap_mode("plot")
tmap_mode("view")

# set bins. right-closed: (a, b] includes b, but NOT a
# that also means that to lump counties with 0, we need to set a value right under
brks <- c(-Inf, -28, -21, -14, -7, -1e-9, Inf)
labs <- c("4 or more weeks lost",
          "3-4 weeks lost",
          "2-3 weeks lost",
          "1-2 weeks lost",
          "Under 1 week lost",
          "Freezing days gained or no change")


# brown color palette
pal <- c("#4A2C1A","#6A3B22","#8A4A2B","#B1704B","#D9B08C","#effeff")

# a few more options
# pal <- c("#00441B","#1A5E2E","#2F7F45","#5AAE61","#A6DDA0","#FFF6E0")
# pal <- c("#7F0000","#B30000","#D7301F","#EF6548","#FCAE91","#F1F7FF")
# pal <- c("#54278f","#6A51A3","#756bb1","#9e9ac8","#cbc9e2","#FFF6E0")

total_change_county_geom %>%
  select(place, freezing_days_change, geometry) %>%
  st_as_sf() %>%
  tm_shape() +
  tm_polygons(col = "freezing_days_change",
              style = "fixed",
              breaks = brks,
              labels = labs,
              palette = pal,
              title = "Change in freezing days",
              border.col = "black",
              lwd = 0.1,
              textNA = "NA") +
  tm_layout(legend.outside = TRUE, frame = FALSE,
    main.title = "Change in freezing days per season, compared to the mid 1950s")

# let's also export data as a geojson file
# we will only keep essential columns and we will smooth geometries
# so that the file is not massive
# total_change_county_geom %>%
#   select(GEOID, county_name, freezing_days_change = freezing_days_change, geometry) %>%
#   st_as_sf() %>%
#   ms_simplify(keep = 0.25, keep_shapes = TRUE) %>% # adjust as needed
#   st_make_valid() %>%
#   st_write("outputs/mapping/total_change_county_v2.geojson",
#          delete_dsn = TRUE)
```